---
title: "Completed Exercises from Hands-On Programming with R"
author: "Gavin McNicol"
date: "1/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

### Acknowledgement

These are the completed exercises for the 10 chapters of *Hands-On Programming with R* authored by Garrett Grolemund. The full text is freely available at <https://rstudio-education.github.io/hopr>.

### Purpose

Although the online version of the textbook includes most of the solutions, they are embedded within all the other material. I therefore created this set of completed exercises  to:

 - **provide a concise template of completed solutions for the coding exercises**
 
But also, to:

 - **practice using pulling and pushing to Github repositories**
 - **gain experience with RMarkdown syntax**

### Getting Started with R

Although I highly recommend the use of [RStudio Cloud](https://rstudio.cloud/), the textbook *Hands-On Programming with R* also features a handy [Appendix](https://rstudio-education.github.io/hopr/starting.html) to guide you through setting up R and RStudio IDE (Integrated Development Environment) on your local computer.

## Chapter 1

### Objects

You can assign an object a name and a value as follows

```{r}
# use <- to assign the value of 1 to the object 'a'
a <- 1
# then test it
a + 2
# the output of the expression evaluated (a + 2) is displayed in the box above
# note the use of the pound sign (#) at the start of a line lets me annotate the code
```

```{r}
# create an object 'die that is a sequence of numbers from 1 to 6, using ':'
die <- 1:6
die
```

R also understands capitalization in object names

```{r}
Name <- 1
name <- 0

name + 1
```

R overwrites previous information without asking permission

```{r}
my_number <- 1
my_number

my_number <- 999
my_number
```

Look at all the objects you have created using 'ls()'
```{r}
ls()
```

You can clear all your work space objects using (remove the # first)

```{r}
# rm(list=ls())
```


You can do all sorts of math with die (a sequence of numeric integers)

```{r}
die - 1
die / 2
die * die
```

When you multiply two vectors of unequal length, R will repeat the shorter vector to match the long vector (known as 'vector recycling')

```{r}
die + 1:2
```


But it will produce an error when the longer objeect is not a multiple of the shorter object

```{r}
die + 1:4
```

You can also do matrix multiplication

```{r}
die %*% die # inner
die %o% die # outer
```

### Functions

### Sample with replacement

### Writing your own functions

### Arguments

### Extract Function in RStudio
